Up to 25\% of beneficial mutations in human protein sequences are not adaptive innovations in mammals

### 2023-01-16 manuscript/main.tex
Updated manuscript, changed colors.

\title{Up to 25\% of beneficial mutations in protein sequences are not adaptive innovations in mammals}

\author{
    \large
    \textbf{T. {Latrille}$^{1}$, J. {Joseph}$^{2}$, N. {Salamin}$^{1}$}\\
    \normalsize
    $^{1}$Université de Lausanne, Lausanne, Switzerland\\
    $^{2}$Université de Lyon, CNRS, LBBE UMR 5558, Villeurbanne, France \\
    \texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}

@@

    \keywords{Adaptation \and Back-mutation \and Phylogenetic \and Population-genetics \and Codon models }

    Adaptation is considered as the main process shaping the diversity of forms and functions across the tree of life~\cite{darwin_origin_1859}.
    It enables species to access new ecological niches and also respond to changes in their environment~\cite{darwin_origin_1859}.
    The notion of adaptation, in its evolutionary sense, is tightly linked to the notion of change, both of the environment, and of the species responding to the latter~\cite{merrell_adaptive_1994}.
    In genomes, adaptation leaves marks of accelerated evolution since beneficial mutations fix faster than neutral ones (fig.~\ref{fig:fitness-landscape}A).
    The signature of an accelerated rate of substitutions can thus be interpreted as a sign of adaptation~\cite{mcdonald_adaptative_1991, smith_adaptive_2002a, welch_estimating_2006}.
    The current availability of large scale genomic data and the development of theoretical models allowed to detect and quantify adaption across genes and lineages, which is now a common practice in evolutionary biology~\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
    Although these approaches led to a better understanding of the processes shaping the rates of molecular evolution, it had the effect to associate beneficial mutations with adaptive evolution, with a loss of the notion of change.
    However, we argue that they should be dissociated since adaptive evolution is not the only process which can lead to beneficial mutations~\cite{charlesworth_other_2007, mustonen_fitness_2009}.
    Indeed, even in a constant environment for which the population is already adapted, a deleterious mutation in an individual can reach fixation in the population by genetic drift~\cite{Ohta1992}.
    Subsequently, a mutation restoring the ancestral fitness will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though no change in the environment occurred~\cite{hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
    More precisely, this restoration of the ancestral fitness can happen at another loci, in which case it is referred as compensating mutations~\cite{hartl_compensatory_1996, mustonen_fitness_2009}.
    Conversely, the restoration of the ancestral fitness can happen at the same locus, in which case it is referred as a beneficial back-mutation~\cite{piganeau_estimating_2003, charlesworth_other_2007}, which are the focus of this study.
    If compensatory mutations still induce changes in the genetic sequence and thus genetic diversification, beneficial-back mutations reduce the diversity and do not contribute to genetic innovation, they cancel the deleterious changes.
    Taken together, not all beneficial mutations contribute to genetic innovation, and truly adaptive evolution can only be well estimated under the background of beneficial back-mutations~\cite{keightley_what_2010, rice_evolutionarily_2015}.
    % Philosophically, our distinction between beneficial back-mutations and adaptation can appear to be far-fetched, and one can argue that beneficial back-mutations is adaptation to its own deteriorating genome.
    % However, there is a difference between adapting to a mutation, for instance by the mean of a compensatory mutation elsewhere, and cancelling the mutation, by the mean of a substitution on the misfit amino-acid.

    For an individual carrying a beneficial mutation, whether this mutation is a new innovation responding to a change in environment or a beneficial back-mutation is indistinguishable~\cite{charlesworth_other_2007}.
    Likewise, at the level of the population, both will result in a positive transmission bias of the beneficial allele.
    However, their consequences at the macro-evolutionary scale are fundamentally different.
    While adaptive evolution promotes phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promote phenotype stability and preserves well established biological systems (fig.~\ref{fig:fitness-landscape}D).
    Fundamentally, the direction of adaptive evolution is unpredictable because it is caused by an unpredictable change in the environment and the underlying fitness landscape~\cite{bazykin_changing_2015}.
    On the other hand, beneficial back-mutations are predictable if the underlying stable fitness landscape is known.
    As a result, a way to quantify the amount of back-mutation is to estimate the underlying fitness landscape, assuming it is constant and stable across time.

    In this study, fitness landscapes of protein-coding genes are estimated from multi-species sequence alignments.
    Multi-species sequence alignments can be used to estimate amino acid fitness, since for example a deleterious amino acid will be more likely to be absent in the alignment.
    One of the most used methods to make these predictions is SIFT (Sort Intolerant From Tolerant), which reliably detects deleterious mutations in many taxa~\cite{ng_sift_2003, vaser_sift_2016}.
    However, SIFT scores is not interpretable as a selection coefficient.
    Moreover, SIFT scores do not acknowledge some key processes that can influence the abundance of an amino-acid in an alignment. For instance, it does not take into account phylogenetic inertia, mutation biases, GC-biased gene conversion, or mutational paths.
    Contrarily, so-called mutation-selection codon models are rooted within a population-genetic formalism~\cite{halpern_evolutionary_1998, mccandlish_modeling_2014}.
    These phylogenetic codon models can disentangle processes of mutation at the DNA level and selection at the amino acid level.
    Such codon models provide a nearly-neutral framework of evolution by estimating the fitness landscape over amino acid sequences for each site of the sequence while taking into account non-adaptive processes impacting codon frequency~\cite{halpern_evolutionary_1998, rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022a}.
    Importantly, mutation-selection codon models assume that the underlying fitness landscape is stable along the phylogeny, hence is a null model of evolution without adaptation.
    From the fitness landscape obtained by fitting the model to the multi-species sequence alignments (fig~\ref{fig:method}A), one can compute the scaled selection coefficient under this null model of evolution, denoted $\Sphy$, of any non-synonymous mutations.
    In this context, a mutation with a positive $\Sphy$ would be driving the current site toward a fitter amino-acid, and thus considered a beneficial back-mutation (fig~\ref{fig:method}B).
    At the genome-wide level, our expectation is that slightly deleterious mutations that are reaching fixations due to drift are scattered across the genome, such that beneficial back-mutations concurrently segregating are also expected to be spread across genes, generating a balance at which genomes are constantly both damaged and repaired simultaneously at different loci.
    Even though each of these back-mutations is expected to have individually a small beneficial effect on their bearer, we expect that the genome-wide signature of beneficial back-mutations can be detected and quantified.

    In this study, by integrating datasets at different scales (phylogeny and population), we seek to predict whether substitutions in terminal lineages or any mutation currently segregating in a population are beneficial back-mutations, i.e with a positive $\Sphy$.
    Such a dataset can allow us to address different questions regarding the selection affecting protein-coding DNA sequences.
    Can we observe beneficial back-mutations that reached fixation recently, and one that are currently segregating in populations?
    Are they effectively beneficial for the individual carrying it?
    Conversely, what is the probability for a beneficial mutation to be a back-mutation?
    In other words, can we find empirical evidence and can we quantify the amount of beneficial mutations that are restoring damaged genomes instead of creating adaptive innovations.
    To answer these questions, we estimated the fitness landscape of amino-acids for long-term evolution using available multiple sequence alignments between mammalian species~\cite{ranwez_orthomam_2007, howe_ensembl_2021}.
    We then integrated data at the phylogenetic and population scale to estimate the amount of beneficial mutations across the entire exome (fig.~\ref{fig:method}A-F), for 28 populations from 6 genera (\textit{Equus},  \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus} and \textit{Homo}).

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.fitness-landscape}
        \caption{
            Panel A and B: at a given site inside a protein-coding DNA sequence, the different amino acids (x-axis) have different fitnesses (y-axis).
            Under a fluctuating fitness landscape (panel A), these fitnesses are changing with time.
            The protein sequence is always lagging behind the moving target defined by the amino acid fitnesses, and since substitutions are accepted preferentially if they are in the direction of this target, substitutions are on average adaptive.
            At the phylogenetic scale (panel C), this phenomena promotes phenotype diversification across species.
            Under a stable fitness landscape (panel B), all the mutations reaching fixation are either slightly deleterious and reaching fixation due to drift, or are beneficial back-mutations restoring a more optimal amino acid.
            At the phylogenetic scale (panel D), this phenomena promotes promotes phenotype stability and preserves well established biological systems.
            At the genome level, slightly deleterious mutations that have reached fixation are supposedly scattered at different loci, such that beneficial back-mutations should also be spread across the genome.
            We thus expect that the genome-wide signature of beneficial back-mutations can be detected and quantified, even though each mutation has individually a small beneficial effect on their bearer.
        }
        \label{fig:fitness-landscape}
    \end{figure*}

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.method.proba}
        \caption{
            Integrating datasets at different scales.
            At the phylogenetic level (panel A), the amino acid Wrightian fitness for each site is estimated from protein-coding DNA alignments using mutation-selection codon models.
            Mutation-selection models assume a stable fitness landscape, taking into account only non-adaptive evolution.
            A mutation with a $\Sphy > 1$ would be driving the current site toward are fitter amino-acid, and such mutation are considered a beneficial back-mutation repairing existing functions (panel B).
            At the population scale (panel C), each single nucleotide polymorphism (SNP) segregating in the population can be classified as a beneficial back-mutation (in red) or not.
            The number of derived allele allow to estimate the probability of beneficial mutations at the population scale, for all SNPs or among back-mutations (panel D).
            The probability of back-mutation is obtained as their fraction among all possible mutations from the ancestral genome sequence (panel E).
            Altogether, we can estimate the probability of back-mutations among all beneficial ones, using Bayes' theorem (panel F).
        }
        \label{fig:method}
    \end{figure*}

    \subsection*{Back-mutations in the terminal lineages}
    We first investigated whether we could find empirical evidence of beneficial back-mutation in the terminal branches of mammals.
    We thus reconstructed the substitutions occurring along the terminal branches of the phylogenetic tree.
    For each substitution, we estimated its scaled selection coefficient ($\Sphy$) based on the stable fitness landscape of amino-acids estimated at the scale of the mammals' phylogeny (fig.~\ref{fig:method}A).
    We showed that among all the substitutions in the terminal branches, between 10 and 13\% had $\Sphy > 1$ (table~S1) and between 25 and 30\% had $\Sphy > 0$ (table~S2).
    For substitutions with $\Sphy > 1$, the ratio of non-synonymous over synonymous divergence, called $\dnds$, was estimated to be between 1.15 and 1.68 in the different lineages (table~S1).
    This meant that beneficial back-mutations with $\Sphy > 1$ reach fixation more frequently than synonymous mutations, which are supposed to be neutral, thus are effectively beneficial.
    This also means that the ratio $\dnds$ usually used to detect adaptation while relying on substitutions between close species~\cite{mcdonald_adaptative_1991, galtier_adaptive_2016} or at the phylogenetic scale~\cite{goldman_codonbased_1994, yang_codonsubstitution_2002} is overinflated due to beneficial back-mutations.
    To alleviate this bias, $\dnds$ is computed exclusively on non-synonymous sites with a $\Sphy < 0$, representing the rate of evolution discarding beneficial back-mutations by construction.
    By dividing $\dnds$ estimated on all substitutions by the one estimated only on substitutions predicted to be deleterious, we directly get the proportion of the rate of evolution which is contributed by beneficial back-mutations (see Material \& Methods).
    This way, we estimated that between 22\% and 27\% of $\dnds$ is inflated because of beneficial back-mutations while they only represent between 1\% and 2\% of mutational opportunities (table~S2).

    \subsection*{Back-mutations in the populations}
    Given that beneficial back-mutations are effectively selected for, we then ask which proportion of new mutations are beneficial back-mutations and are also beneficial at the population scale (fig.~\ref{fig:method}C-E).
    For each population with available polymorphism data, we first reconstructed the ancestral genome (instead of the reference genome) at the base of the population genealogy.
    From this reference sequence, we computed the scaled selection coefficient ($\Sphy$) for each possible non-synonymous mutations, weighted by their mutational opportunity (see Material \& Methods).
    We thus obtained the distribution of fitness effects (DFE) of new mutations from the ancestral sequence in \textit{Homo sapiens} (fig.~\ref{fig:homo-afr-results}A), under the assumption of the stable fitness landscape.
    Because $\Sphy$ is based on estimation of fitnesses at the mammalian scale, mutations with $\Sphy>1$ are considered beneficial back-mutations.
    Beneficial back-mutations only represents \~1.5\% of all possible new mutations (red in fig.~\ref{fig:homo-afr-results}A, table~S3).
    However, when focusing on the mutations that are observed in populations, namely single nucleotide polymorphism (SNP), beneficial back-mutations represents \~8.5\% of all observed mutations (red in fig.~\ref{fig:homo-afr-results}B, table~S3).

    Additionally, in populations, the frequencies at which SNPs are segregating is informative of their selective effect, e.g. deleterious SNPs are segregating at lower frequencies because they are purified.
    As such, methods have been developed to estimate the distribution of fitness effect (DFE) at the population scale by gathering information across many SNPs~\cite{eyre-walker_distribution_2006a, eyre-walker_estimating_2009a, galtier_adaptive_2016, tataru_inference_2017}.
    These methods offer a unique opportunity to compare selection coefficients estimated at the population scale ($\Spop$) and at the phylogenetic scale ($\Sphy$).
    In order to estimate the DFE at the population level and gather robust statistical estimates, we grouped SNPs observed at the population level (e.g.~in humans of African descent in fig.~\ref{fig:homo-afr-results}) in 5 classes of predicted selection coefficient: severely deleterious with $\divStrongDel$, deleterious with $\divDel$, weakly deleterious with $\divWeakDel$, weakly beneficial with $\divWeakAdv$ and beneficial with $\divAdv$ (fig.~\ref{fig:homo-afr-results}B).

    To obtain a quantitative estimate of the distribution of selection coefficient for each category of SNPs, we used the model polyDFE~\cite{tataru_inference_2017, tataru_polydfe_2020}.
    This model uses the count of derived alleles for each frequency class (fig.~\ref{fig:homo-afr-results}C) to infer the distribution of fitness effects (DFE).
    Moreover, the model relies on the mutational opportunities ($L$) for each category of $\Sphy$ (fig.~\ref{fig:homo-afr-results}A, see Material \& Methods) as well as the count of synonymous derived alleles used as a neutral model, which allow to control for demographic effects, and polarization errors when inferring the ancestral states.
    We could estimate the proportion of beneficial mutations ($\PpolyAdv$), nearly-neutral mutations ($\PpolyNeutral$) and deleterious mutations ($\PpolyDel$) for each class of selection previously defined (fig.~\ref{fig:homo-afr-results}D).
    We found that SNPs predicted to be deleterious based on the selection coefficient $\Sphy$ estimated across the phylogeny of mammals were effectively purged at the population scale (blue and green columns in fig.~\ref{fig:homo-afr-results}D)
    Purifying selection is therefore largely predictable and amino acids with low fitness across mammals are effectively purged away in each population.
    Then, SNPs predicted to be beneficial based on the selection coefficient $\Sphy$ estimated at the scale of the phylogeny of mammals are indeed beneficial for individuals currently bearing them (red column in fig.~\ref{fig:homo-afr-results}D).
    These results confirmed that selection towards amino acids that are restoring existing function is ongoing in these populations.
    Between these two extreme cases, SNPs predicted to be weakly selected (cyan and orange columns in fig.~\ref{fig:homo-afr-results}D) were effectively composed of neutral, beneficial and deleterious mutations.

    We replicated our experiment across 28 populations of different mammal species (fig.~\ref{fig:heatmap}) and found that SNPs predicted to be deleterious were indeed purged at the population scale (blue and green rows in fig.~\ref{fig:heatmap}A).
    However, SNPs predicted to be weakly beneficial ($\divWeakAdv$) were composed of a mix of neutral and selected mutations with varying proportions across the different populations (cyan and orange row in fig.~\ref{fig:heatmap}B).
    Finally, SNPs predicted to be beneficial back-mutations ($\divAdv$) were found to be beneficial with varying proportions across populations (red row in fig.~\ref{fig:heatmap}C).
    The variable proportions found between populations can be explained by the effective number of individuals in the population ($\Ne$), which is well-known to be a major driver of selection efficacy.
    Since $\Ne$ is not directly measurable, we used the Watterson $\theta_w$ as a proxy to order populations by their levels of synonymous diversity.
    Higher synonymous diversity is associated with a smaller proportion of nearly-neutral mutations (fig.~\ref{fig:diversity}A\&B).
    This result confirms the nearly-neutral theory and suggests that populations with higher diversity (e.g.~\textit{Bos} or \textit{Ovis}) are more likely to discriminate between beneficial and deleterious mutations (fig~S1-S9), while mutations in populations with low diversity (e.g.~\textit{Homo}) are effectively nearly-neutral.

>>>>

\title{Up to 25\% of beneficial mutations in human protein sequences are not adaptive innovations in mammals}

\author{
    \large
    \textbf{T. {Latrille}$^{1}$, J. {Joseph}$^{2}$, D.~A. {Hartasánchez}$^{1}$, N. {Salamin}$^{1}$}\\
    \normalsize
    $^{1}$Université de Lausanne, Lausanne, Switzerland\\
    $^{2}$Université de Lyon, CNRS, LBBE UMR 5558, Villeurbanne, France \\
    \texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}

@@

    \keywords{Adaptation \and Back-mutation \and Phylogenetics \and Population-genetics \and Codon models }

    Adaptation is considered one of the main processes shaping the diversity of forms and functions across the tree of life~\cite{darwin_origin_1859}.
    It enables species to access new ecological niches and also respond to changes in their environment~\cite{darwin_origin_1859} [NS: maybe cite Losos or some key papers on ecological speciation rather than Darwin twice].
    In evolution, adaptation is tightly linked to the notion of change at the level of the environment and to species responding to this change~\cite{merrell_adaptive_1994}.
    The process of adaptation leaves marks of accelerated evolution in genomes since mutations that are beneficial under new conditions fix faster than neutral mutations (fig.~\ref{fig:fitness-landscape}A).
    The signature of an accelerated rate of substitutions, defined as the rate of mutation times the probability of fixation, can thus be interpreted as a sign of adaptation~\cite{mcdonald_adaptative_1991, smith_adaptive_2002, welch_estimating_2006}.
    The current availability of large-scale genomic data and the development of theoretical models have allowed to detect and quantify adaptation across genes and lineages, and these analyses have become common practice in evolutionary biology~\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019} [NS: cite the paper by Anna Marcionetti here].
    Although these approaches have led to a better understanding of the processes determining the rates of molecular evolution, they have had the effect of associating beneficial mutations with adaptive evolution, while disconnecting them from the notion of change defining adaptation.
    We argue that we should limit the use of adaptive mutations to those that are associated with adaptation as such and dissociate beneficial mutations from adaptive mutations since adaptive evolution is not the only process that can lead to beneficial mutations~\cite{charlesworth_other_2007, mustonen_fitness_2009}.
    We propose here a new approach to identify the mechanisms underlying the origin of beneficial mutations.
    In particular, we suggest that many beneficial mutations are not adaptive, but are rather restoring ancestral states of higher fitness.
    We used large-scale genomic data to integrate analyses at both the phylogenetic and population scales to differentiate between truly adaptive mutations from mutations restoring the ancestral fitness.
    Our results show that up to a quarter of beneficial mutations across mammal species are not driven by adaptive innovations, but rather compensate the effect of ancestral mutations to recover the optimal state of the species.

    In a constant environment, if a population is already adapted to these conditions, a deleterious mutation in an individual can reach fixation in the population by genetic drift~\cite{Ohta1992}.
    Subsequently, a new mutation that is restoring the ancestral fitness will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though no change in the environment occurred~\cite{hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
    The restoration of the ancestral fitness can happen at a different locus than the initial mutation, in which case it is referred to as a compensating [NS: ou compensatory ?] mutation~\cite{hartl_compensatory_1996, mustonen_fitness_2009}.
    However, if the restoration of the ancestral fitness happens at the same locus than the initial mutation, in which case it is referred to as a beneficial back-mutation~\cite{piganeau_estimating_2003, charlesworth_other_2007}.
    In this study, we focus on the latter case.
    While compensatory mutations induce genetic changes at the sequence level, and thus genetic diversification, beneficial back-mutations reduce the genetic diversity and do not contribute to genetic innovation.
    They merely cancel the effect of the previous deleterious changes.
    As a consequence, not all beneficial mutations contribute to genetic innovation and truly adaptive evolution can only be correctly estimated if we can account for the number of beneficial back-mutations~\cite{keightley_what_2010, rice_evolutionarily_2015}.
    % Philosophically, our distinction between beneficial back-mutations and adaptation can appear to be far-fetched, and one can argue that beneficial back-mutations is adaptation to its own deteriorating genome.
    % However, there is a difference between adapting to a mutation, for instance by means of a compensatory mutation elsewhere, and cancelling the mutation, by means of a substitution on the misfit amino-acid.

    %For an individual carrying a beneficial mutation, whether this mutation is a new innovation responding to a change in environment or a beneficial back-mutation is indistinguishable~\cite{charlesworth_other_2007}.
    Given a beneficial mutation, distinguishing whether it is a new innovation responding to a change in environment or a beneficial back-mutation is impossible with only one individual at hand~\cite{charlesworth_other_2007}.
    This limitation persists at the level of the population, since both types of mutation will result in a positive transmission bias of the beneficial allele.
    However, at the macro-evolutionary scale the consequences of these two types of mutations are fundamentally different.
    While adaptive evolution promotes phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promote phenotype stability and preserve well established biological systems (fig.~\ref{fig:fitness-landscape}D).
    Additionally, the direction of adaptive evolution is unpredictable because it is caused by an unpredictable change in the environment and, hence, in the underlying fitness landscape~\cite{bazykin_changing_2015}.
    On the other hand, beneficial back-mutations are predictable if the underlying stable fitness landscape is known because we will expect to see changes from non-optimal amino acids to optimal ones.
    If we wish to quantify the amount of beneficial back-mutations that have happened along a phylogeny we need to identify the selection coefficients of mutations and to achieve this we require the fitness landscape to be stable and be known.

    The placental mammals represent an excellent study system to perform this analysis.
    Having originated ~100 Mya\cite{kumar_timetree_2017}, they diversified quickly.
    Additionally, we have polymorphism data available for many species\cite{howe_ensembl_2021} and high quality protein-coding DNA alignments across the genome\cite{ranwez_orthomam_2007, scornavacca_orthomam_2019}.
    By restricting our analysis to protein-coding orthologous genes to all the phylogeny, and further excluding any genes for which there have been claims of pervasive directional selection, we can not only assume a constant fitness landscape for each of them throughout placental mammal evolution, but actually reconstruct the fitness landscape from variable positions within the sequence along the phylogeny.
    For each gene, we used a codon model of mutation-selection~\cite{halpern_evolutionary_1998, mccandlish_modeling_2014} to obtain relative fitnesses for all amino-acids for each site of the sequence by fitting the model to the multi-species sequence alignment (fig~\ref{fig:method}A) and assuming that the underlying fitness landscape is stable along the phylogenetic tree.
    %Mutation-selection codon models provide a nearly neutral framework of evolution taking into account non-adaptive processes impacting codon frequency~\cite{halpern_evolutionary_1998, rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022}.
    From the fitness landscape, we can calculate the scaled selection coefficient $\Sphy$ for any possible amino-acid change at each position.
    Under this model, a mutation with a positive $\Sphy$ would be driving the current site toward a known fitter amino-acid and would thus be considered as a beneficial back-mutation (fig~\ref{fig:method}B), given the assumed stability of the fitness landscape.

    Having identified which potential amino-acid changes could represent beneficial back-mutations at the phylogenetic scale, we retrieved polymorphism data from 28 populations belonging to 6 genera (\textit{Equus}, \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus} and \textit{Homo}) to assess the presence of beneficial back-mutations at the population scale.
    We focused on currently segregating mutations within populations and also on substitutions in the terminal lineages, and checked if any of these observed changes where predicted beneficial back-mutations.
    Essentially, by integrating large scale genomic datasets at both phylogenetic and population scales we have been able to estimate the amount of beneficial back-mutations across the entire exome for 6 different mammalian genera (fig~\ref{fig:method}A-F).

    Our analyses allow us to test long-standing questions in molecular evolution:
    How common are beneficial back-mutations?
    Can we identify beneficial back-mutations that have reached fixation recently, and ones that are currently segregating in populations?
    Are hypothetical beneficial back-mutations effectively beneficial for the individuals carrying them?
    Furthermore, what is the probability for any beneficial mutation observed in current populations to be a back-mutation?
    In other words, can we quantify the amount of beneficial mutations that are restoring damaged genomes instead of creating adaptive innovations.
    By addressing these questions we provide
    novel insight on how selection is affecting protein-coding DNA sequences.

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.fitness-landscape}
        \caption{
            Panel A and B: at a given site of a protein-coding DNA sequence, the different amino acids (x-axis) have different fitnesses (y-axis).
            Under a fluctuating fitness landscape (panel A), these fitnesses are changing with time.
            The protein sequence is always lagging behind the moving target defined by the amino acid fitnesses, and since substitutions are accepted preferentially if they are in the direction of this target, substitutions are on average adaptive.
            At the phylogenetic scale (panel C), this phenomena promotes phenotype diversification across species.
            Under a stable fitness landscape (panel B), all the mutations reaching fixation are either slightly deleterious and reaching fixation due to drift, or are beneficial back-mutations restoring a more optimal amino acid.
            At the phylogenetic scale (panel D), this phenomena promotes promotes phenotype stability and preserves well established biological systems.
            At the genome level, slightly deleterious mutations that have reached fixation are supposedly scattered at different loci, such that beneficial back-mutations should also be spread across the genome.
            We thus expect that the genome-wide signature of beneficial back-mutations can be detected and quantified, even though each mutation has individually a small beneficial effect on their bearer.
        }
        \label{fig:fitness-landscape}
    \end{figure*}

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.method.proba}
        \caption{
            Integrating datasets at different scales.
            At the phylogenetic level (panel A), the amino acid Wrightian fitness for each site is estimated from protein-coding DNA alignments using mutation-selection codon models.
            Mutation-selection models assume a stable fitness landscape, taking into account only non-adaptive evolution.
            A mutation with a $\Sphy > 1$ would be driving the current site toward are fitter amino-acid, and such mutation are considered a beneficial back-mutation repairing existing functions (panel B).
            At the population scale (panel C), each single nucleotide polymorphism (SNP) segregating in the population can be classified as a beneficial back-mutation (in blue) or not.
            The number of derived alleles allows us to estimate the probability of beneficial mutations at the population scale, for all SNPs or among back-mutations (panel D).
            The probability of back-mutation is obtained as their fraction among all possible mutations from the ancestral genome sequence (panel E).
            Altogether, we can estimate the probability of back-mutations among all beneficial ones, using Bayes' theorem (panel F).
        }
        \label{fig:method}
    \end{figure*}

    \subsection*{Back-mutations in the terminal lineages}
    We first estimated the number of beneficial back-mutation in the terminal branches of mammals by reconstructing the substitutions occurring along the terminal branches of the phylogenetic tree.
    For each substitution, we estimated its scaled selection coefficient ($\Sphy$) based on the stable fitness landscape of amino-acids estimated at the scale of the phylogenetic tree of mammals (fig.~\ref{fig:method}A).
    Among all the substitutions that were found in the terminal branches, between 10 and 13\% had $\Sphy > 1$ (table~S1), while between 25 and 30\% had $\Sphy > 0$ (table~S2).
    For substitutions with $\Sphy > 1$, the $\dnds$ ratio of non-synonymous over synonymous divergence was estimated to be between 1.15 and 1.68 in the different lineages (table~S1).
    Beneficial back-mutations with $\Sphy > 1$ therefore reached fixation more frequently than synonymous mutations that are supposed to be neutral, which means that these back-mutations are effectively beneficial.
    This result further indicated that $\dnds$ used to detect adaptation while relying on substitutions between closely related species~\cite{mcdonald_adaptative_1991, galtier_adaptive_2016} or at the phylogenetic scale~\cite{goldman_codonbased_1994, yang_codonsubstitution_2002} is biased due to the presence of beneficial back-mutations that are not adaptive.
    We can alleviate this bias by computing $\dnds$ exclusively on non-synonymous sites with a $\Sphy < 0$, which represents the rate of evolution when beneficial back-mutations are discarded.
    By comparing these two ways of estimating $\dnds$ (see Material \& Methods), we estimated that between 22\% and 27\% of $\dnds$ is inflated because of beneficial back-mutations while they only represent between 1\% and 2\% of mutational opportunities (table~S2).

    \subsection*{Back-mutations in the populations}
    We then estimated the proportion of new mutations that were beneficial back-mutations while being under selection at the population scale (fig.~\ref{fig:method}C-E).
    We first reconstructed the ancestral genome (instead of the reference genome) of each mammal species with available polymorphism data.
    We computed the scaled selection coefficient ($\Sphy$) along this ancestral genome for each possible non-synonymous mutations, weighted by their mutational opportunity (see Material \& Methods).
    It allowed us to obtain the distribution of fitness effects (DFE) of new mutations from the ancestral sequence under the assumption of the stable fitness landscape as shown for \textit{Homo sapiens} in Fig.~\ref{fig:homo-afr-results}A.
    Because $\Sphy$ was based on estimation of fitnesses at the mammalian scale, mutations with $\Sphy>1$ were considered, as shown above, beneficial back-mutations and only represented ca.~1.5\% of all possible new mutations (blue in fig.~\ref{fig:homo-afr-results}A, table~S3).
    However, when focusing on the mutations that are observed in populations, namely single nucleotide polymorphism (SNP), beneficial back-mutations represents ca.~8.5\% of all observed mutations (blue in fig.~\ref{fig:homo-afr-results}B, table~S3).

    Within populations, the frequencies at which SNPs are segregating is also providing information on their selective effect.
    For instance, deleterious SNPs are segregating at lower frequencies because of the effect of purifying selection, which removes them from the population.
    It is possible to estimate the DFE at the population scale by gathering information across many SNPs~\cite{eyre-walker_distribution_2006, eyre-walker_estimating_2009, galtier_adaptive_2016, tataru_inference_2017} and these approaches offer a unique opportunity to compare selection coefficients estimated at the population scale ($\Spop$) and at the phylogenetic scale ($\Sphy$).
    We estimated the DFE at the population level and gathered robust statistical estimates by grouping SNPs observed in 5 classes of predicted selection coefficient (e.g.~in humans of African descent in fig.~\ref{fig:homo-afr-results}): severely deleterious with $\divStrongDel$, deleterious with $\divDel$, weakly deleterious with $\divWeakDel$, weakly beneficial with $\divWeakAdv$ and beneficial with $\divAdv$ (fig.~\ref{fig:homo-afr-results}B).
	%in mat/met
	%To obtain a quantitative estimate of the distribution of selection coefficient for each category of SNPs, we used the model polyDFE~\cite{tataru_inference_2017, tataru_polydfe_2020}.
    %This model uses the count of derived alleles for each frequency class (fig.~\ref{fig:homo-afr-results}C) to infer the distribution of fitness effects (DFE).
    %Moreover, the model relies on the mutational opportunities ($L$) for each category of $\Sphy$ (fig.~\ref{fig:homo-afr-results}A, see Material \& Methods) as well as the count of synonymous derived alleles used as a neutral model, which allow to control for demographic effects, and polarization errors when inferring the ancestral states.
    From the DFE, we further estimated the proportion of beneficial mutations ($\PpolyAdv$), nearly-neutral mutations ($\PpolyNeutral$) and deleterious mutations ($\PpolyDel$) for each class of selection previously defined (fig.~\ref{fig:homo-afr-results}D).
    We found that SNPs predicted to be deleterious based on the selection coefficient $\Sphy$ estimated across the phylogenetic tree of mammals were effectively purged at the population scale (red and yellow columns in fig.~\ref{fig:homo-afr-results}D)
    Purifying selection was therefore largely predictable and amino acids with low fitness across mammals were effectively purged away in each population.
    Then, SNPs predicted to be beneficial based on the selection coefficient $\Sphy$ estimated at the phylogenetic scale were indeed beneficial for individuals currently bearing them (blue column in fig.~\ref{fig:homo-afr-results}D).
    These results confirmed that selection towards amino acids that are restoring existing function is ongoing in these populations.
    Between these two extreme cases, SNPs predicted to be weakly selected (cyan and orange columns in fig.~\ref{fig:homo-afr-results}D) were effectively composed of neutral, beneficial and deleterious mutations.

    We replicated our experiment across 28 populations of different mammal species (fig.~\ref{fig:heatmap}) and found that SNPs predicted to be deleterious were indeed purged at the population scale (red and yellow rows in fig.~\ref{fig:heatmap}A).
    However, SNPs predicted to be weakly beneficial ($\divWeakAdv$) were composed of a mix of neutral and selected mutations with varying proportions across the different populations (cyan and orange row in fig.~\ref{fig:heatmap}B).
    Finally, SNPs predicted to be beneficial back-mutations ($\divAdv$) were found to be beneficial with varying proportions across populations (blue row in fig.~\ref{fig:heatmap}C).
    The variable proportions found between populations can be explained by the effective number of individuals in the population ($\Ne$), which is known to be a major driver of selection efficacy.
    Since $\Ne$ is not directly measurable, we used the Watterson $\theta_w$ as a proxy to order populations by their levels of synonymous diversity.
    Higher synonymous diversity was associated with a smaller proportion of nearly-neutral mutations (fig.~\ref{fig:diversity}A\&B).
    This result confirmed the nearly-neutral theory and suggested that populations with higher diversity (e.g.~\textit{Bos} or \textit{Ovis}) were more likely to discriminate between beneficial and deleterious mutations (fig~S1-S9), while mutations in populations with low diversity (e.g.~\textit{Homo}) were effectively nearly-neutral.
